\bf An observational study of papers published in \emph{PLOS ONE} and studies posted to a trial registry.

### 2021-05-24 manuscript/asa_template/asa_template.Rmd
update paper add plos flow diagram

All _PLOS ONE_ articles are freely accessible via the PLOS Application Programming Interface (API). This enabled us to conduct searches of full-text articles and analyse data on articles' text content and general attributes such as publication date and field(s) of research. To find papers with a statistical methods section we used targeted API searches followed by article filtering based on section headings. The data were downloaded on 3 July 2020.

_Step 1_: Targeted API searches. API searches were completed using the R package ‘rplos’ [@rplos]. Search queries targeted the presence of analysis-related terms anywhere in the article. Search terms combined the words “data” or “statistical” with one of: “analysis”, “analyses”, “method”, “methodology” or “model(l)ing”. By allowing terms to appear anywhere in the article, we accounted for the possibility of relevant text being placed in different sections, for example, in the _Material and Methods_ section versus _Results_. Search results were indexed by a unique Digital Object Identifier (DOI). Attribute data collected per DOI included journal volume and subject classification(s).

_Step 2_: Partial matching on section headings. Full text XML data were downloaded and combined into a single dataset, organised by DOI and subsection heading(s). Since _PLOS ONE_ does not prescribe standardised headings to preface statistical methods sections, we performed partial matching on available headings against frequently used terms in initial search results: ‘Statistical analysis’, ‘Statistical analyses’, ‘Statistical method’, ‘Statistics’, ‘Data analysis’ and ‘Data analyses’.

<!-- For records that did not pass this second stage filter, we selected a random sample of XXX records and reviewed where initial search terms appeared in the full-text, to estimate the likely proportion of statistical methods sections that were missed. -->

@@

Both datasets were cleaned using functions from the `textclean package' [@textclean]. 
Text cleaning aimed to standardise notation and statistical terminology, whilst minimising changes to article style and formatting. _R_ code used for data extraction and cleaning is available from https://github.com/agbarnett/stats_section.

Mathematical notation including Greek letters was converted from Unicode characters to plain text. For example, the Unicode character corresponding to $\theta$ (<U+03B8>) was replaced with ‘theta’. Common symbols outside of Unicode blocks including ‘%’ (percent) and ‘<’ (‘less-than’) were similarly converted into plain text. General formatting was removed, this included carriage returns, punctuation marks, in-text references (e.g. "[42]") centred equations, and other non-ASCII characters. Text contained inside brackets was retained to maximise content for analysis, with brackets removed.

We compiled an extensive list of statistical terms to standardise descriptions of statistical methods reported across both datasets. An initial list was compiled by calculating individual word frequencies and identifying relevant terms that appeared at least 100 times. Further terms were sourced from index searches of three statistics textbooks [@Dobson2018,@Diggle2013,@Bland2015]. The final list is provided in Supplementary File 1. Plurals (e.g., ‘chi-squares’) unhyphenated (e.g., ‘chi square’) and combined (e.g. ‘chisquare’) terms were transformed to singular, hyphenated form (e.g., ‘chi-square’). Common statistical tests were also hyphenated (e.g., ‘hosmer lemeshow’ to ‘hosmer-lemeshow’).

As a final step, common stop words including pronouns, contractions and selected prepositions were removed. We retained selected stop words that, if excluded, may have changed the context of statistical methods being described, for example ‘between’ and ‘against’.

<!-- Questions: -->

<!-- * Reference updates to stats reporting guidelines; e.g see: -->

<!-- - https://everyone.plos.org/2019/09/26/new-plos-one-statistical-reporting-guidelines/. -->
<!-- - https://web.archive.org/web/20190607174803/https://journals.plos.org/plosone/s/submission-guidelines -->
<!-- - https://web.archive.org/web/20150507175314/https://journals.plos.org/plosone/s/submission-guidelines -->

@@

Let $P \in R^{M \times N}$ denotes the paper content matrix where the  statistical methods sections of $N$ papers consist of $M$ distinct terms. Text clustering for identifying common writing themes in these papers requires this matrix to represent with a vector space model. Let Matrix $P$ be modeled with the unique terms in the processed sections represented with the tf\*idf (term frequent * inverse document frequency) weighting schema to consider both common and rare terms.

Text clustering is known to face the curse of dimensionality due to the high number of terms in doc$\times$term matrix representation [@aggarwal2012mining, @sutanto2018fine]. Therefore text-based methods based on distance, density or probability face difficulties [@park2018examining,@mohotti2018corpus,@mohotti2019concept]. Specifically, the distance difference between near and far points becomes negligible in high-dimensional data [@aggarwal2012mining]. This directly affects the distance-based methods such as \textit{k}-means \cite{jain} in accurately identifying the common subgroups. In addition, the sparseness of this high dimensional matrix representation does not allow to differentiate the user groups based on density differences [@mohotti2018corpus, @mohotti2018efficient].

Non-negative Matrix factorization (NMF) which maps the high-dimensional data to a lower-dimensional space has been found to provide an effective solution by allowing to form clusters in the lower-dimensional space [@kim2014algorithms,@luong2019clustering]. NMF approximates topic vectors in linear manner considering the context of terms.

In traditional NMF [@aggarwal2012mining], the high-dimensional matrix \( P \in R^{M \times N} \) is approximated by learning two factor matrices \( W \in R^{M \times g} \) and \( H \in R ^{N \times g} \)  where $g$ is the number of cluster groups or common topics in the data collection, as follows.

\begin{equation}
\label{eq:2}
 P \approx WH^{T}
\end{equation}

The matrix factorization process approximates the lower dimensional non-negative factor matrices $W$ and $H$ such that they can represent high dimensional $P$ with the least error. The objective function in Eq \ref{eq:Eq2} iteratively attempts to update the entries in $W$ and $H$ to find the optimum values which possess minimum sum of square error for all the elements in both of those matrices. NMF uses Frobenius norm as its objective function and will find the optimum value of \textit{W} and \textit{H} iteratively.

\begin{equation}
\label{eq:Eq2}
\min \frac{1}{2}\|P - WH\|= \sum _{i=1}^{M}\sum _{j=1}^{N} \left(  P_{i,j} -\left(WH \right)_{i,j} \right)^{2}
\end{equation}

Matrix \textit{H} contains the information regarding topic membership of each document. Topic membership of each method section in $P$ is obtained considering the maximum coefficient value in $H$ for a method section.

We applied the NMF clustering algorithm to the cleaned dataset, varying the number of clusters from 1 to 50. Clustering solutions were assessed using silhouette score, within-cluster dispersion and between-cluster
dispersion.

@@

\subsection{Missing statistical methods sections}

@@

## _PLOS ONE_

@@

knitr::include_graphics("figures/plos.summary.jpg")
```

@@

API searches returned `r format(n.api,big.mark=',')` results (DOIs) based on chosen search terms. After partial matching, `r format(n.include,big.mark=',')` (`r perc.include`%) statistical methods sections were identified. In the final sample, 95,518 (85%) DOIs returned an exact match against common section headings: 64,133 for 'statistical analysis', 13,380 for 'statistical analyses' and 13,627 for 'data analysis'. For our random sample of DOIs that did not meet the partial matching criteria, initial search terms appeared in Introduction (number), Results (number) only.  xxx appeared in the Methods section without a section heading and xxx had a non-standard section heading. [TODO].

@@

Statistical methods sections had a median length of `r n.words[['med']]` words and inter-quartile range of `r n.words[['q1']]` to `r n.words[['q3']]` words (Figure \ref{fig:plos-n}B). `r format(n.extremes[['morethan500']],big.mark=',')` articles (`r p.extremes[['morethan500']]`%) had a statistical methods section of 500 words or more. `r format(n.extremes[['lessthan50']],big.mark=',')` articles (`r p.extremes[['lessthan50']]`%) had sections with 50 words or less, equal to the length of this paragraph.

All DOIs included Biology and life sciences (n = `r format(n.biology,big.mark=',')`), Earth sciences
(n = `r format(n.earth,big.mark=',')`) and/or Computer and information sciences (n = `r format(n.computer,big.mark=',')`) in their top 3 subject classifications (Figure \ref{fig:plos-n}C).

@@

knitr::include_graphics("figures/plos.wordclouds.jpg")

@@

knitr::include_graphics("figures/anzctr.wordclouds.jpg")

>>>>

All _PLOS ONE_ articles are freely accessible via the PLOS Application Programming Interface (API). This enabled us to conduct searches of full-text articles and analyse data on articles' text content and general attributes such as publication date and field(s) of research. Statistical methods sections were extracted using a two-step approach:

_Step 1_: Targeted API searches were completed using the R package ‘rplos’ [@rplos]. Search queries targeted analysis-related terms, combining the words “data” or “statistical” with one of: “analysis”, “analyses”, “method”, “methodology” or “model(l)ing”. Terms could appear anywhere within the main body of the article, to account for the placement of relevant text in different sections, for example, in the _Material and Methods_ section versus _Results_. Search results were indexed by a unique Digital Object Identifier (DOI). Attribute data collected per DOI included journal volume and subject classification(s).

_Step 2_: _PLOS ONE_ does not prescribe standardised headings to preface statistical methods sections. To address this, we performed partial matching on available headings against frequently used terms in initial search results: ‘Statistical analysis’, ‘Statistical analyses’, ‘Statistical method’, ‘Statistics’, ‘Data analysis’ and ‘Data analyses’.

<!-- Full-text XML data were combined into a single dataset organised by DOI and subsection heading(s).  -->

Data included in our analysis were downloaded on 3 July 2020. For records that did not pass Step 2, we reviewed where initial search terms appeared in the full-text, to evaluate the likely proportion of statistical methods sections that were missed.

@@

<!-- Datasets were cleaned using functions from the `textclean package' [@textclean].  -->

Text cleaning aimed to standardise notation and statistical terminology, whilst minimising changes to article style and formatting. _R_ code used for data extraction and cleaning is available from https://github.com/agbarnett/stats_section.

Mathematical notation was converted from Unicode characters to plain text. For example, the Unicode character corresponding to $\theta$ (<U+03B8>) was replaced with ‘theta’. Common symbols outside of Unicode blocks including ‘%’ (percent) and ‘<’ (‘less-than’) were similarly converted into plain text. General formatting was removed, including carriage returns, punctuation marks, in-text references (e.g. "[42]") centred equations, and other non-ASCII characters. Text contained inside brackets was retained to maximise content for analysis, with brackets removed. Common stop words including pronouns, contractions and selected prepositions were removed. We retained selected stop words that, if excluded, may have changed the context of statistical methods being described, for example ‘between’ and ‘against’.

We compiled an extensive list of statistical terms to standardise reported descriptions of statistical methods. An initial list was compiled by calculating individual word frequencies and identifying relevant terms that appeared at least 100 times. Further terms were sourced from index searches of three statistics textbooks [@Dobson2018,@Diggle2013,@Bland2015]. Plurals (e.g., ‘chi-squares’) unhyphenated (e.g., ‘chi square’) and combined (e.g. ‘chisquare’) terms were transformed to singular, hyphenated form (e.g., ‘chi-square’). Common statistical tests were also hyphenated (e.g., ‘hosmer lemeshow’ to ‘hosmer-lemeshow’). The final list is provided in Supplementary File 1.

@@

Text from statistical methods sections was analysed using Non-Negative Matrix Factorization (NMF). NMF is an established approach for topic modelling, and provides an effective solution for text-based clustering when dealing with high-dimensional data [@kim2014algorithms,@luong2019clustering]. In this section, we outline key details of NMF and its strengths compared with other text-based clustering methods.

For $N$ studies, let $P \in R^{M \times N}$ denote a content matrix, representing text from statistical methods sections as $M$ unique terms. Text clustering algorithms for identifying common topics across studies requires $P$ to be represented with a vector space model. In our case, unique terms in $P$ are modelled using the tf\*idf (term frequency $\times$ inverse document frequency) weighting schema, to account for the relative importance of common and rare terms.

A common problem facing text clustering algorithms is the curse of dimensionality due to the high number of terms in the doc$\times$term matrix representation [@aggarwal2012mining, @sutanto2018fine]. Applying text-based methods based on distance, density or probability therefore face difficulties in high-dimensional settings [@park2018examining,@mohotti2018corpus,@mohotti2019concept]. Specifically, distances between near and far points becomes negligible [@aggarwal2012mining]. This behaviour directly affects the performance of distance-based clustering methods such as \textit{k}-means [@jain] in accurately identifying subgroups (topics) present in the data. Furthermore, sparseness associated with high-dimensional matrix representations does not allow for differentiation between topics based on density differences [@mohotti2018corpus, @mohotti2018efficient].

To address these limitations, NMF deals with high-dimensional data by mapping it to a lower-dimensional space. This mapping this achieved by approximating $P$ with two factor matrices: \( W \in R^{M \times g} \) and \( H \in R ^{N \times g} \) [@aggarwal2012mining]. The number of subgroups of common topic inferred from the data is represented by $g$, leading to the approximation:
\begin{equation}
\label{eq:approxPNMF}
 P \approx WH^{T}.
\end{equation}

The matrix factorization process approximates the lower dimensional non-negative factor matrices $W$ and $H$ such that they can represent high dimensional $P$ with the least error. Estimation of $W$ and $H$ is achieved by optimising an objective function; for NMB, the Fronbeius norm is used, equivalent to minimising the sum of squares for all elements of $P$:

\begin{equation}
\label{eq:NMFobjectivefn}
\min \frac{1}{2}\|P - WH\|= \sum _{i=1}^{M}\sum _{j=1}^{N} \left(  P_{i,j} -\left(WH \right)_{i,j} \right)^{2}
\end{equation}

Following estimation, \textit{H} contains the information regarding topic membership for all studies. In our case, topic membership $(1,\ldots,g)$ for a statistical methods section is inferred from the maximum coefficient value in the corresponding row of $H$.

For our two datasets, we applied NMF varying $g$ from 1 to 50. Clustering solutions were assessed using silhouette score, within-cluster dispersion and between-cluster dispersion.

@@

<!-- \subsection{Missing statistical methods sections} -->

@@

## Included studies

### PLOS ONE

@@

knitr::include_graphics("figures/plos_summary.jpg")
```

@@

Searching with the _PLOS ONE_ API returned `r format(n.api,big.mark=',')` results (DOIs) based on chosen search terms (\autoref{fig:plos-flowchart}). After partial matching, `r format(n.include,big.mark=',')` (`r perc.include`%) statistical methods sections were identified. In the final sample, 95,518 (85%) DOIs returned an exact match against common section headings: 64,133 for 'statistical analysis', 13,380 for 'statistical analyses' and 13,627 for 'data analysis'. For our random sample of DOIs that did not meet the partial matching criteria, initial search terms appeared in Introduction (number), Results (number) only.  xxx appeared in the Methods section without a section heading and xxx had a non-standard section heading. [TODO].

@@

All DOIs included Biology and life sciences (n = `r format(n.biology,big.mark=',')`), Earth sciences
(n = `r format(n.earth,big.mark=',')`) and/or Computer and information sciences (n = `r format(n.computer,big.mark=',')`) in their top 3 subject classifications (Figure \ref{fig:plos-n}B).

Statistical methods sections had a median length of `r n.words[['med']]` words and inter-quartile range of `r n.words[['q1']]` to `r n.words[['q3']]` words. `r format(n.extremes[['morethan500']],big.mark=',')` articles (`r p.extremes[['morethan500']]`%) had a statistical methods section of 500 words or more. `r format(n.extremes[['lessthan50']],big.mark=',')` articles (`r p.extremes[['lessthan50']]`%) had statistical methods sections with 50 words or less, equal to the length of this paragraph.

@@

knitr::include_graphics("figures/ploswordclouds.jpg")

@@

knitr::include_graphics("figures/anzctrwordclouds.jpg")

### 2021-05-24 manuscript/asa_template/asa_template.tex
update paper add plos flow diagram

All \emph{PLOS ONE} articles are freely accessible via the PLOS
Application Programming Interface (API). This enabled us to conduct
searches of full-text articles and analyse data on articles' text
content and general attributes such as publication date and field(s) of
research. To find papers with a statistical methods section we used
targeted API searches followed by article filtering based on section
headings. The data were downloaded on 3 July 2020.

\emph{Step 1}: Targeted API searches. API searches were completed using
the R package `rplos' \citep{rplos}. Search queries targeted the
presence of analysis-related terms anywhere in the article. Search terms
combined the words ``data'' or ``statistical'' with one of:
``analysis'', ``analyses'', ``method'', ``methodology'' or
``model(l)ing''. Search terms were intended to be broad whilst keeping
search results to a manageable number for full-text review (see Step 2).
By allowing terms to appear anywhere in the article, we accounted for
the possibility of relevant text being placed in different sections, for
example, in the \emph{Material and Methods} section versus
\emph{Results}. Search results were indexed by a unique Digital Object
Identifier (DOI). Attribute data collected per DOI included journal
volume and subject classification(s).

\emph{Step 2}: Partial matching on section headings. Full text XML data
for all search results were downloaded and combined into a single
dataset, organised by DOI and subsection heading(s). Since \emph{PLOS
ONE} does not prescribe standardised headings to preface statistical
methods sections, we performed partial matching on available headings
against frequently used terms in initial search results: `Statistical
analysis', `Statistical analyses', `Statistical method', `Statistics',
`Data analysis' and `Data analyses'. For records that did not pass this
second stage filter, we selected a random sample of XXX records and
reviewed where initial search terms appeared in the full-text, to
estimate the likely proportion of statistical methods sections that were
missed.

@@

We applied the same text cleaning steps to both data sources using
functions from the `textclean package' \citep{textclean}. Text cleaning
aimed to standardise notation and statistical terminology, whilst
minimising changes to article style and formatting. \emph{R} code used
for data extraction and cleaning is available from
\url{https://github.com/agbarnett/stats_section}.

Mathematical notation including Greek letters was converted from Unicode
characters to plain text. For example, the Unicode character
corresponding to \(\theta\) (\textless U+03B8\textgreater) was replaced
with `theta'. Common symbols outside of Unicode blocks including `\%'
(percent) and `\textless{}' (`less-than') were similarly converted into
plain text. General formatting was removed, this included carriage
returns, punctuation marks, in-text references (e.g.~``{[}42{]}'')
centred equations, and other non-ASCII characters. Text contained inside
brackets was retained to maximise content for analysis, with brackets
removed.

We compiled an extensive list of statistical terms to standardise
descriptions of statistical methods reported across both datasets. An
initial list was compiled by calculating individual word frequencies and
identifying relevant terms that appeared at least 100 times. Further
terms were sourced from index searches of three statistics textbooks
\citep[\citet{Diggle2013},\citet{Bland2015}]{Dobson2018}. The final list
is provided in Supplementary File 1. Plurals (e.g., `chi-squares')
unhyphenated (e.g., `chi square') and combined (e.g.~`chisquare') terms
were transformed to singular, hyphenated form (e.g., `chi-square').
Common statistical tests were also hyphenated (e.g., `hosmer lemeshow'
to `hosmer-lemeshow').

As a final step, common stop words including pronouns, contractions and
selected prepositions were removed. We retained selected stop words
that, if excluded, may have changed the context of statistical methods
being described, for example `between' and `against'.

@@

\emph{Details to come}

We applied the clustering algorithm to the cleaned dataset, varying the
number of clusters from 1 to 50. Clustering solutions were assessed
using silhouette score, within-cluster dispersion and between-cluster
dispersion.

For each dataset, records assigned to invididual clusters were examined
for evidence of boilerplate text in two ways. We first reviewed the
strongest matches to each cluster based on topic scores{[}may not be the
right word{]}. We also compared similarities in text between pairs of
records assigned to each cluster. Similarities were determined from
calculating cosine similarity scores, with comparisons limited to the
top 500 records per cluster. Higher similarity score represent greater
overlap in text content between statistical methods sections. Results
were transformed to lower case for the clustering, but examples are
given using the original capitalisation.

\subsection{Missing statistical methods sections}

@@

\hypertarget{plos-one}{%
\subsection{\texorpdfstring{\emph{PLOS ONE}}{PLOS ONE}}\label{plos-one}}

@@

{\centering \includegraphics[width=0.6\linewidth]{figures/plos.summary}

@@

API searches returned 131,847 results (DOIs) based on chosen search
terms. After partial matching, 111,731 (85\%) statistical methods
sections were identified. In the final sample, 95,518 (85\%) DOIs
returned an exact match against common section headings: 64,133 for
`statistical analysis', 13,380 for `statistical analyses' and 13,627 for
`data analysis'. For our random sample of DOIs that did not meet the
partial matching criteria, initial search terms appeared in Introduction
(number), Results (number) only. xxx appeared in the Methods section
without a section heading and xxx had a non-standard section heading.
{[}TODO{]}.

@@

Statistical methods sections had a median length of 127 words and
inter-quartile range of 61 to 254 words (Figure \ref{fig:plos-n}B).
7,450 articles (7\%) had a statistical methods section of 500 words or
more. 19,461 articles (17\%) had sections with 50 words or less, equal
to the length of this paragraph.

All DOIs included Biology and life sciences (n = 107,584), Earth
sciences (n = 7,605) and/or Computer and information sciences (n =
5,190) in their top 3 subject classifications (Figure
\ref{fig:plos-n}C).

@@

{\centering \includegraphics[width=0.8\linewidth]{figures/plos.wordclouds}

@@

\begin{landscape}
\begin{table}[]
\centering
\caption{Example boilerplate text from PLOS ONE dataset}
\label{tab:plos-example-boilerplate}
\begin{tabular}{p{0.1\linewidth} p{0.9\linewidth}}
\hline
\textbf{Topic} & \textbf{Statistical methods section} \\ \hline
1 & Student’s t-test was used for statistical analysis. A p value of $<$0.05 was considered statistically significant. \\
& Statistical analysis was performed using the Students t-test. P values of less than 0.05 were considered significant.\\
\hline
3 & GraphPad Prism (Graphpad Software, San Diego, CA) was used for all analyses.\\
 & All statistical analysis was performed using Graphpad Prism software.\\
 \hline
4 & Significant differences were determined using analysis of variance (ANOVA) followed by Tukey post-hoc tests for multiple comparisons.\\
& Data are expressed as the mean ± SEM. Statistical analysis was performed using one-way analysis of variance (ANOVA) test, followed by Dunnett’s multiple comparison post hoc test. P values less than 0.05 were considered significant.\\
\hline
5 & SPSS software version 17.0 (SPSS, Chicago, IL, USA) was used for statistical analysis.\\
  & The results are presented as the mean ± SEM. Data were analyzed by one-way ANOVA and LSD tests using the SPSS 13.0 software (SPSS Inc., Chicago, IL, USA). The difference was considered statistically significant at P$<$0.05.\\
  \hline
6 & All results are expressed as means ± standard deviation (± SD).\\
& Data are expressed as mean ± Standard error of the mean (SEM). Statistical analysis of the data was performed by the Student’s t-test. A value of P$<$0.05 was considered statistically significant.\\
\hline
9 & Statistical significance was determined by Students t-tests. $^{*} P<0.05$, $^{**} P = <0.01$, $^{***} P<0.001$.\\
 & Data are presented as the mean ± SD. Statistical analysis was performed using the Students unpaired t-test. Differences were considered statistically significant at $^{*}p<0.05$, $^{**}p<0.01$ and $^{***}p<0.001$.\\
 \hline
\end{tabular}
\end{table}
\end{landscape}

@@

>>>>

All \emph{PLOS ONE} articles are freely accessible via the PLOS
Application Programming Interface (API). This enabled us to conduct
searches of full-text articles and analyse data on articles' text
content and general attributes such as publication date and field(s) of
research. Statistical methods sections were extracted using a two-step
approach:

\emph{Step 1}: Targeted API searches were completed using the R package
`rplos' \citep{rplos}. Search queries targeted analysis-related terms,
combining the words ``data'' or ``statistical'' with one of:
``analysis'', ``analyses'', ``method'', ``methodology'' or
``model(l)ing''. Terms could appear anywhere within the main body of the
article, to account for the placement of relevant text in different
sections, for example, in the \emph{Material and Methods} section versus
\emph{Results}. Search results were indexed by a unique Digital Object
Identifier (DOI). Attribute data collected per DOI included journal
volume and subject classification(s).

\emph{Step 2}: \emph{PLOS ONE} does not prescribe standardised headings
to preface statistical methods sections. To address this, we performed
partial matching on available headings against frequently used terms in
initial search results: `Statistical analysis', `Statistical analyses',
`Statistical method', `Statistics', `Data analysis' and `Data analyses'.

Data included in our analysis were downloaded on 3 July 2020. For
records that did not pass Step 2, we reviewed where initial search terms
appeared in the full-text, to evaluate the likely proportion of
statistical methods sections that were missed.

@@

Text cleaning aimed to standardise notation and statistical terminology,
whilst minimising changes to article style and formatting. \emph{R} code
used for data extraction and cleaning is available from
\url{https://github.com/agbarnett/stats_section}.

Mathematical notation was converted from Unicode characters to plain
text. For example, the Unicode character corresponding to \(\theta\)
(\textless U+03B8\textgreater) was replaced with `theta'. Common symbols
outside of Unicode blocks including `\%' (percent) and `\textless{}'
(`less-than') were similarly converted into plain text. General
formatting was removed, including carriage returns, punctuation marks,
in-text references (e.g.~``{[}42{]}'') centred equations, and other
non-ASCII characters. Text contained inside brackets was retained to
maximise content for analysis, with brackets removed. Common stop words
including pronouns, contractions and selected prepositions were removed.
We retained selected stop words that, if excluded, may have changed the
context of statistical methods being described, for example `between'
and `against'.

We compiled an extensive list of statistical terms to standardise
reported descriptions of statistical methods. An initial list was
compiled by calculating individual word frequencies and identifying
relevant terms that appeared at least 100 times. Further terms were
sourced from index searches of three statistics textbooks
\citep[\citet{Diggle2013},\citet{Bland2015}]{Dobson2018}. Plurals (e.g.,
`chi-squares') unhyphenated (e.g., `chi square') and combined
(e.g.~`chisquare') terms were transformed to singular, hyphenated form
(e.g., `chi-square'). Common statistical tests were also hyphenated
(e.g., `hosmer lemeshow' to `hosmer-lemeshow'). The final list is
provided in Supplementary File 1.

@@

Text from statistical methods sections was analysed using Non-Negative
Matrix Factorization (NMF). NMF is an established approach for topic
modelling, and provides an effective solution for text-based clustering
when dealing with high-dimensional data
\citep[\citet{luong2019clustering}]{kim2014algorithms}. In this section,
we outline key details of NMF and its strengths compared with other
text-based clustering methods.

For \(N\) studies, let \(P \in R^{M \times N}\) denote a content matrix,
representing text from statistical methods sections as \(M\) unique
terms. Text clustering algorithms for identifying common topics across
studies requires \(P\) to be represented with a vector space model. In
our case, unique terms in \(P\) are modelled using the tf*idf (term
frequency \(\times\) inverse document frequency) weighting schema, to
account for the relative importance of common and rare terms.

A common problem facing text clustering algorithms is the curse of
dimensionality due to the high number of terms in the doc\(\times\)term
matrix representation
\citep[\citet{sutanto2018fine}]{aggarwal2012mining}. Applying text-based
methods based on distance, density or probability therefore face
difficulties in high-dimensional settings
\citep[\citet{mohotti2018corpus},\citet{mohotti2019concept}]{park2018examining}.
Specifically, distances between near and far points becomes negligible
\citep{aggarwal2012mining}. This behaviour directly affects the
performance of distance-based clustering methods such as
\textit{k}-means \citep{jain} in accurately identifying subgroups
(topics) present in the data. Furthermore, sparseness associated with
high-dimensional matrix representations does not allow for
differentiation between topics based on density differences
\citep[\citet{mohotti2018efficient}]{mohotti2018corpus}.

To address these limitations, NMF deals with high-dimensional data by
mapping it to a lower-dimensional space. This mapping this achieved by
approximating \(P\) with two factor matrices: \(W \in R^{M \times g}\)
and \(H \in R ^{N \times g}\) \citep{aggarwal2012mining}. The number of
subgroups of common topic inferred from the data is represented by
\(g\), leading to the approximation: \begin{equation}
\label{eq:approxPNMF}
 P \approx WH^{T}.
\end{equation}

The matrix factorization process approximates the lower dimensional
non-negative factor matrices \(W\) and \(H\) such that they can
represent high dimensional \(P\) with the least error. Estimation of
\(W\) and \(H\) is achieved by optimising an objective function; for
NMB, the Fronbeius norm is used, equivalent to minimising the sum of
squares for all elements of \(P\):

\begin{equation}
\label{eq:NMFobjectivefn}
\min \frac{1}{2}\|P - WH\|= \sum _{i=1}^{M}\sum _{j=1}^{N} \left(  P_{i,j} -\left(WH \right)_{i,j} \right)^{2}
\end{equation}

Following estimation, \textit{H} contains the information regarding
topic membership for all studies. In our case, topic membership
\((1,\ldots,g)\) for a statistical methods section is inferred from the
maximum coefficient value in the corresponding row of \(H\).

For our two datasets, we applied NMF varying \(g\) from 1 to 50.
Clustering solutions were assessed using silhouette score,
within-cluster dispersion and between-cluster dispersion.

\subsection{Within topic analysis}

Within the \emph{PLOS ONE} dataset, statistical methods sections with
high topic coherence scores were used to investigate evidence of
boilerplate text, compared with other studies assigned to the same
topic. Comparisons were carried out by calculating Jaccard similarity
scores between studies at the sentence level. We chose the Jaccard score
as an intuitive measure, which summarises the similarity between two
sentences by the number of words common to each sentence (intersection),
divided by the number of words that appeared in either the strongest
match or the sentence being compared. Boilerplate text was defined based
on word count (plus or minus 3 words) and a Jaccard score of 0.9 or
higher.

Results were transformed to lower case for the clustering, but examples
are given using the original capitalisation.

@@

\hypertarget{included-studies}{%
\subsection{Included studies}\label{included-studies}}

\hypertarget{plos-one}{%
\subsubsection{PLOS ONE}\label{plos-one}}

@@

{\centering \includegraphics[width=0.6\linewidth]{figures/plos_summary}

@@

Searching with the \emph{PLOS ONE} API returned 131,847 results (DOIs)
based on chosen search terms (\autoref{fig:plos-flowchart}). After
partial matching, 111,731 (85\%) statistical methods sections were
identified. In the final sample, 95,518 (85\%) DOIs returned an exact
match against common section headings: 64,133 for `statistical
analysis', 13,380 for `statistical analyses' and 13,627 for `data
analysis'. For our random sample of DOIs that did not meet the partial
matching criteria, initial search terms appeared in Introduction
(number), Results (number) only. xxx appeared in the Methods section
without a section heading and xxx had a non-standard section heading.
{[}TODO{]}.

@@

All DOIs included Biology and life sciences (n = 107,584), Earth
sciences (n = 7,605) and/or Computer and information sciences (n =
5,190) in their top 3 subject classifications (Figure
\ref{fig:plos-n}B).

Statistical methods sections had a median length of 127 words and
inter-quartile range of 61 to 254 words. 7,450 articles (7\%) had a
statistical methods section of 500 words or more. 19,461 articles (17\%)
had statistical methods sections with 50 words or less, equal to the
length of this paragraph.

@@

{\centering \includegraphics[width=0.8\linewidth]{figures/ploswordclouds}

@@

\begin{table}

\caption{\label{tab:unnamed-chunk-7}Example boilerplate text from PLOS ONE dataset}
\centering
\begin{tabular}[t]{l|l|r}
\hline
Topic & Statistical methods text & Number of matching studies\\
\hline
Topic 1 & a p value of less-than 0.05 was considered statistically significant. & 1883\\
\hline
Topic 1 & p values of less-than 0.05 were considered significant. & 1526\\
\hline
Topic 1 & statistical analysis was performed using the students t-test & 380\\
\hline
Topic 1 & students t-test was used for statistical analysis & 29\\
\hline
Topic 3 & graphpad prism graphpad software san diego ca was used for all analyses. & 1941\\
\hline
Topic 3 & all statistical analysis was performed using graphpad prism software. & 1312\\
\hline
Topic 4 & statistical analysis was performed using one-way analysis of variance anova test followed by dunnetts multiple comparison post hoc test & 4516\\
\hline
Topic 4 & p values less-than 0.05 were considered significant. & 2414\\
\hline
Topic 4 & significant differences were determined using analysis of variance anova followed by tukey post-hoc tests for multiple comparisons. & 2216\\
\hline
Topic 4 & data are expressed as the mean plus-or-minus sem & 860\\
\hline
Topic 5 & the difference was considered statistically significant at p less-than 0.05. & 2555\\
\hline
Topic 5 & spss software version 17.0 spss chicago il usa was used for statistical analysis. & 441\\
\hline
Topic 5 & the results are presented as the mean plus-or-minus sem & 420\\
\hline
Topic 5 & data were analyzed by one-way anova and lsd tests using the spss 13.0 software spss inc chicago il usa & 19\\
\hline
Topic 6 & a value of p less-than 0.05 was considered statistically significant. & 1786\\
\hline
Topic 6 & data are expressed as mean plus-or-minus standard error of the mean sem & 604\\
\hline
Topic 6 & statistical analysis of the data was performed by the students t-test & 444\\
\hline
Topic 6 & all results are expressed as means plus-or-minus standard deviation plus-or-minus sd. & 102\\
\hline
Topic 9 & statistical analysis was performed using the students unpaired t-test & 752\\
\hline
Topic 9 & differences were considered statistically significant at p less-than 0.05 p less-than 0.01 and p less-than 0.001. & 413\\
\hline
Topic 9 & data are presented as the mean plus-or-minus sd & 378\\
\hline
Topic 9 & statistical significance was determined by students t-tests & 49\\
\hline
Topic 9 & p less-than 0.05 p equal-to less-than 0.01 p less-than 0.001. & 6\\
\hline
\end{tabular}
\end{table}

@@

Finally, targeted searches for the term `statistician' revealed
instances across all topics where no methods were specified. Instead,
all analyses were the responsibility of an appointed statistician, for
example, ``A statistician employed by hospital was used'' and ``Pilot
study at this point will use a statistician professionally to determine
sample size calculations as required''.
